---
title: "Assessing Canada’s Drug Shortage Problem"
#author: "Jon Pipitone"
output: 
  flexdashboard::flex_dashboard:
    source_code: embed
---

```{r setup}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse, warn.conflicts = FALSE)
library(ggrepel)
library(lubridate)

# Terminology 
# DPD = Drug Product Database
# DSC = drugshortagescanada.ca - the current, mandatory reporting shortage databsae
# DS  = drugshortages.ca - the previous, voluntary reporting shortage datbase

# Settings (note: some figures have settings in their chunks)
DSC_CSV = 'data/drugshortagescanada.ca-2017-12-31.csv'
DS_MANUAL_CSV = 'data/drugshortages.ca-manual.csv'
DS_EXPORT_CSV = 'data/drugshortages.ca-export.csv'
ANALYSIS_PERIOD_START = as.Date("2013-04-01")   # start of manual collection period
ANALYSIS_PERIOD_END   = as.Date("2016-03-31")   # end of manual collection period

DPD_YEAR_START = ymd('1980-01-01') # earliest year to look at in the DPD
DPD_YEAR_END = ymd('2017-01-01')   # latest year to look at in the DPD 

# set to TRUE to attempt to download the latest DPD dataset
DPD_DOWNLOAD = FALSE 

# URL for the drug product database files
DPD_URL_BASE='http://www.hc-sc.gc.ca/dhp-mps/alt_formats/zip/prodpharma/databasdon/'
```

```{r load-drug-shortage-data}
# New Drug Shortages database
drugshortagescanada_ca_raw = read_csv(DSC_CSV, col_types=cols(
  DIN=col_character(),
  `Shortage Start Date` = col_date(), 
  `Anticipated Shortage Start Date` = col_date(), 
  `Actual Shortage End Date` = col_date(), 
  `Estimated Shortage End Date` = col_date()
  )) 
drugshortagescanada_ca = drugshortagescanada_ca_raw %>% 
  mutate(start_date = if_else(is.na(`Shortage Start Date`), 
                              `Anticipated Shortage Start Date`, `Shortage Start Date`),
         end_date = if_else(is.na(`Actual Shortage End Date`), 
                              `Estimated Shortage End Date`, `Actual Shortage End Date`),
         source = 'drugshortagescanada.ca', 
         resolution = 'Current') %>%
  select(DIN, start_date, end_date, source, Reason, resolution)


# Manually collected drugshortages.ca data collected during the analysis period
drugshortages_ca_manual_raw = read_csv(DS_MANUAL_CSV)
drugshortages_ca_manual = drugshortages_ca_manual_raw %>%
  transmute(
    DIN = din, 
    start_date = as.Date(start_date), 
    end_date = as.Date(est_resupply_date),
    source = 'drugshortages.ca-manual', 
    Reason = 'Unknown', 
    resolution = outcome, 
    resolution = 'unknown'
  )

# Export of the drugshortages.ca database obtained from the maintainer
drugshortages_ca_export_raw = read_csv(DS_EXPORT_CSV)
drugshortages_ca_export = drugshortages_ca_export_raw %>% 
  transmute(
    DIN = str_pad(DIN, 8, pad = "0"),
    start_date = as.Date(`Start Date`),
    end_date = as.Date(`Estimated Resupply Date`),
    source = "drugshortages.ca-export", 
    Reason = `Reason for Shortage`, 
    resolution = if_else(is.na(Discontinued), Status, Discontinued))
```

```{r merge-shortage-datasets}
# merge the drugshortages.ca datasets
# We will only include data from the Export after the end of the analysis period
drugshortages_ca = drugshortages_ca_export %>%
  filter(start_date > ANALYSIS_PERIOD_END) %>%
  bind_rows(drugshortages_ca_manual) %>%
  group_by(DIN, start_date) %>%                   # remove duplicates
  summarize(
    end_date = first(end_date), 
    source = 'drugshortages.ca',
    Reason = first(Reason), 
    resolution = first(resolution)) %>%
  ungroup()

# merge the datasets
shortages = bind_rows(drugshortagescanada_ca, drugshortages_ca) %>% 
  group_by(DIN, start_date) %>%
  summarize(source = if_else(n() == 1, paste(source, collapse=" "), 'both')) %>%
  ungroup()
```

```{r load-dpd-dataset}
# Load the Health Canada Drug Product Database
DPD_DATA_DIR = 'data/dpd/'

if(DPD_DOWNLOAD) {
  for (f in c('allfiles.zip', 'allfiles_ia.zip', 'allfiles_ap.zip', 'allfiles_dr.zip')) {
    destfile = paste0(DPD_DATA_DIR, f)
    download.file(paste0(DPD_URL_BASE, f), destfile)
    unzip(destfile, exdir = DPD_DATA_DIR)
  }
}

tables = list(
  comp = c("DRUG_CODE","MFR_CODE","COMPANY_CODE","COMPANY_NAME","COMPANY_TYPE",
           "ADDRESS_MAILING_FLAG","ADDRESS_BILLING_FLAG",
           "ADDRESS_NOTIFICATION_FLAG","ADDRESS_OTHER","SUITE_NUMBER",
           "STREET_NAME","CITY_NAME","PROVINCE","COUNTRY","POSTAL_CODE",
           "POST_OFFICE_BOX","PROVINCE_F","COUNTRY_F"),
  form = c("DRUG_CODE","PHARM_FORM_CODE","PHARMACEUTICAL_FORM",
           "PHARMACEUTICAL_FORM_F"),
  ingred = c("DRUG_CODE","ACTIVE_INGREDIENT_CODE","INGREDIENT",
             "INGREDIENT_SUPPLIED_IND","STRENGTH","STRENGTH_UNIT",
             "STRENGTH_TYPE","DOSAGE_VALUE","BASE","DOSAGE_UNIT","NOTES",
             "INGREDIENT_F","STRENGTH_UNIT_F","STRENGTH_TYPE_F","DOSAGE_UNIT_F"),
  package = c("DRUG_CODE","UPC","PACKAGE_SIZE_UNIT","PACKAGE_TYPE",
              "PACKAGE_SIZE","PRODUCT_INFORMATION","PACKAGE_SIZE_UNIT_F",
              "PACKAGE_TYPE_F"),
  pharm = c("DRUG_CODE","PHARMACEUTICAL_STD"),
  drug = c("DRUG_CODE","PRODUCT_CATEGORIZATION","CLASS",
           "DRUG_IDENTIFICATION_NUMBER","BRAND_NAME","DESCRIPTOR",
           "PEDIATRIC_FLAG","ACCESSION_NUMBER","NUMBER_OF_AIS",
           "LAST_UPDATE_DATE","AI_GROUP_NO","CLASS_F","BRAND_NAME_F",
           "DESCRIPTOR_F"),
  route = c("DRUG_CODE","ROUTE_OF_ADMINISTRATION_CODE",
            "ROUTE_OF_ADMINISTRATION","ROUTE_OF_ADMINISTRATION_F"),
  sched = c("DRUG_CODE","SCHEDULE","SCHEDULE_F"),
  status = c("DRUG_CODE","CURRENT_STATUS_FLAG","STATUS","HISTORY_DATE",
             "STATUS_F","LOT_NUMBER","EXPIRATION_DATE"),
  ther = c("DRUG_CODE","TC_ATC_NUMBER","TC_ATC","TC_AHFS_NUMBER","TC_AHFS",
           "TC_ATC_F","TC_AHFS_F"),
  vet = c("DRUG_CODE","VET_SPECIES","VET_SUB_SPECIES","VET_SPECIES_F")
)

for (table_name in names(tables)) {
  headers = tables[[table_name]]
  # create new variables based on the DPD table names (e.g. dpd_vet) and load in the
  # data from the corresponding files (active, inactive, etc..) and merge them
  assign(paste0('dpd_', table_name), 
    dir(DPD_DATA_DIR, pattern = paste0(table_name, '.*.txt$'), full.names = TRUE) %>%
      keep(function(x) file.size(x) > 0) %>%
      map(read_csv, 
          col_names = headers, 
          locale = locale('en', date_format = '%d-%b-%Y'), 
          col_types = map(headers,  function (x) # guess base on name 
            case_when(
              endsWith(x, '_DATE') ~ 'D', 
              x == 'DRUG_CODE' ~ 'i', 
              TRUE ~ 'c')) %>% 
          setNames(headers), progress = FALSE) %>%
      reduce(rbind)
  )
}

# Restrict ourselves to just the prescription drugs for humans
drug_product = dpd_drug %>%
  filter(CLASS == 'Human') %>%
  select(DRUG_CODE, DRUG_IDENTIFICATION_NUMBER, AI_GROUP_NO) %>%
  mutate(IG = substr(AI_GROUP_NO, 3, 7))

drug_codes_human = select(drug_product, DRUG_CODE)

drug_sched_prescription = dpd_sched %>% 
  filter(SCHEDULE == 'Prescription')

drug_status = dpd_status %>% 
  mutate(HISTORY_DATE = as.Date(HISTORY_DATE),
         STATUS = fct_shift(as_factor(STATUS),2)) %>%
  inner_join(drug_codes_human, by = 'DRUG_CODE') %>%
  inner_join(drug_sched_prescription, by = 'DRUG_CODE') 

# make a list of years
years = tibble(
    year.date = seq(DPD_YEAR_START, DPD_YEAR_END, by='years'), 
    n = 1)

# compute drug statuses for each year
# We do this with a full cross join, and then filter non-sense rows
# (An hack unworthy of a true computer scientist... but definitely worth a try)
drug_yearly_status = drug_status %>%
  mutate(n=1) %>%
  full_join(years, by = 'n') %>%
  filter(HISTORY_DATE <= year.date) %>%
  group_by(year.date, DRUG_CODE) %>%
  arrange(HISTORY_DATE, .by_group = T) %>%
  summarize(status = last(STATUS)) 

```

Drug Shortages
==============

Row {data-width=300}
--------------------

### Shortage Datasets
We have two sources of data on drug shortages: 

#### drugshortages.ca
- Active until March, 2017 
- Voluntary reporting
- We manually collected data from this database from 2013-04-01 through 2016-03-31 (`drugshortages.ca-manual.csv`)
- We also obtained exports of this database as of various dates (`drughsortages.ca-export.csv`)

#### drugshortagescanada.ca
- Active from March 2017 to present
- Mandatory reporting
- Exports provided on demand via [online search](https://www.drugshortagescanada.ca/search?term=&date_range%5Bdate_range_start%5D%5Bmonth%5D=&date_range%5Bdate_range_start%5D%5Bday%5D=&date_range%5Bdate_range_start%5D%5Byear%5D=&date_range%5Bdate_range_end%5D%5Bmonth%5D=&date_range%5Bdate_range_end%5D%5Bday%5D=&date_range%5Bdate_range_end%5D%5Byear%5D=&filter_type=shortages&filter_status=_all_)
- Note: The *anticipated shortage start date* is used if the start date is
  blank. If a shortage for the same DIN and start date appears in both datasets it
  is included once.

### Ingredient Groups
Ingredient Groups are taken from the [Drug Product Database](https://www.canada.ca/en/health-canada/services/drugs-health-products/drug-products/drug-product-database/terminology.html) 
active ingredient group number which identifies the unique groups of active
ingredients(s) regardless of strength. This is explained in more detail on the
"Drug Products" tab.

Row {.tabset}
----------------------------------------------------
### Figure 1 - Drug Shortages Per Month
```{r figure-1-din-shortages}

# period of time to display 
YEAR_START = 2014
YEAR_END = 2017

figure_1_data = shortages %>%
  filter(between(year(start_date), YEAR_START, YEAR_END)) %>%
  mutate(start = floor_date(start_date, unit="month"), 
         month = factor(format(start_date, "%b"), levels=month.abb))
  
write_csv(figure_1_data, 'figure_1-shortages-by-start-date.csv')

figure_1 = figure_1_data %>%
  ggplot(aes(x = month, fill = source)) + 
  geom_bar() + 
  labs(y = "# of DINs", x = "Anticipated/Actual Shortage Start Date") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, size=8, vjust=0.3), 
        panel.grid = element_blank(), 
        strip.placement = "outside", 
        legend.position = "bottom") +
  facet_grid(. ~ year(start_date), switch="x") +
  scale_fill_discrete(guide = guide_legend(title = "Source")) + 
  scale_y_continuous(expand=c(0,0)) + 
  labs(
    title = "Drug Shortages per Month, Canada, 2012–16", 
    caption = paste("Last updated:", now()), 
    subtitle = "Duplicates determined by DIN and start date only (and are only counted once)")

ggsave('figure_1-shortages-by-start-date.png', plot = figure_1)
figure_1
```

### Figure 3 - Shortages by Innovator Status
```{r shortages-innovator}
figure_3_data = drugshortages_ca_manual_raw %>% 
  mutate(start_date = as.Date(start_date)) %>%
  mutate(month = factor(format(start_date, "%b"), levels=month.abb), 
         company_type = fct_rev(company_type)) %>%
  filter(between(start_date, ANALYSIS_PERIOD_START, ANALYSIS_PERIOD_END)) 

figure_3_data %>%
  mutate(Year = year(start_date)) %>%
  group_by(Year, month, company_type) %>%
    summarize(count = n()) %>%
  write_csv("figure-3_shortages_by_company_type_by_month.csv")

figure_3 = figure_3_data %>%
  ggplot(aes(x = month, fill = company_type)) +
  #ggplot(aes(x = year(start_date), fill = company_type)) +
    geom_bar() + 
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 90, size=8, vjust=0.3), 
          panel.grid = element_blank(), 
          strip.placement = "outside", 
          legend.position = "bottom") +
    scale_fill_discrete(guide = guide_legend(title = "")) + 
    labs(title = "Shortages of Generic or Innovator Drug Products over a Three-Year Period, Canada, 2013–16", 
         subtitle = "Three year period 2013*-2016**", 
         y = "", 
         x = "Month") + 
  facet_grid( ~ year(start_date))

ggsave('figure-3_shortages-by-company-type.png', plot = figure_3)
figure_3
```

### Figure 5 - Shortages by Ingredient Group

```{r ig-shortages}
TOP_N_IG = 20

# Date range to display shortage ingredient groups
IG_START_DATE = as.Date('2017-03-01') 
IG_END_DATE = as.Date('2017-12-31')

shortages_with_IG = 
  inner_join(drug_product, shortages,  by=c('DRUG_IDENTIFICATION_NUMBER' = 'DIN')) %>%
  inner_join(dpd_route, by='DRUG_CODE') %>%
  filter(ROUTE_OF_ADMINISTRATION == 'ORAL' | ROUTE_OF_ADMINISTRATION == 'INTRAMUSCULAR')

figure_5_data = shortages_with_IG %>%
  mutate(year = year(start_date)) %>%
  filter(between(start_date, IG_START_DATE, IG_END_DATE)) %>%
  group_by(year, IG) %>%
    summarize(n = n()) %>%
    ungroup() %>%
  filter(n > 1) %>%
  arrange(year, desc(n)) %>%
  group_by(year) %>%
    top_n(TOP_N_IG, wt = n) %>%
  ungroup() %>%
  inner_join(drug_product, by = "IG") %>%
  inner_join(dpd_ingred, by = "DRUG_CODE") %>%
  select(IG, n, year, INGREDIENT) %>%
  mutate(INGREDIENT = gsub(' \\(.*\\)','', INGREDIENT)) %>%
  mutate(IG_NAME = paste(INGREDIENT, '(', IG, ')')) %>%
  unique()

write_csv(figure_5_data, 'figure_5-shortages_by_IG.csv')

figure_5 = figure_5_data %>%
  ggplot(aes(x = reorder(IG_NAME, n), y = n, fill = reorder(IG_NAME, n))) +
    theme(axis.text.x = element_text(angle = 90, size=8, vjust=0.3), 
          legend.text=element_text(size=5), 
          legend.position = 'none') +  
    geom_bar(stat = 'identity', na.rm = T) +
    coord_flip() + 
    #facet_grid(. ~ year, scales = "free") + 
    labs(
      title = "Top 21 Shortages by Ingredient (and Group Number), Canada, March–December 2017", 
      subtitle = paste0("Top ", TOP_N_IG, "ish most common Ingredient Groups shown. IM/PO only"), 
      caption = paste('Sources: DPD, DSC. Updated:', now()), 
      x = "Ingredient (Group #)", 
      y = "Number of Shortages"
    ) + 
    scale_fill_discrete(guide = guide_legend(title = "Ingredient (Ingredient Group)")) 

ggsave('figure_5-shortages_by_IG.png', plot = figure_5)
figure_5
```
### Figure 9: Top 10 Drugs MFRs by Shortages

```{r marketed-drugs-in-shortage-by-manufacturer}
# The analysis period starts and ends midway through a year, but we have
# have drugs active per year, so we need to floor our dates. 
MFR_START_YEAR = floor_date(ANALYSIS_PERIOD_START, unit = "year")
MFR_END_YEAR = floor_date(ANALYSIS_PERIOD_END, unit = 'year')
TOP_N_MFR_MARKETED = 10

mfr_products = drug_yearly_status %>%
  filter(between(year.date, MFR_START_YEAR, MFR_END_YEAR), status %in% c('MARKETED')) %>%
  group_by(DRUG_CODE) %>%
  summarise() %>%
  inner_join(dpd_comp, by="DRUG_CODE") %>%
  group_by(COMPANY_NAME) %>%
  summarize(n_products = n())
  
# compute the shortages during the analysis period
mfr_shortages = drugshortages_ca %>% 
  filter(between(start_date, ANALYSIS_PERIOD_START, ANALYSIS_PERIOD_END)) %>%
  inner_join(drug_product, by=c('DIN' = 'DRUG_IDENTIFICATION_NUMBER')) %>%
  inner_join(dpd_comp, by='DRUG_CODE') %>%
  group_by(COMPANY_NAME) %>%
  summarize(n_shortages = n())

# join and compute the percentage shortages of marketed drugs
mfr_percent_shortages = inner_join(mfr_shortages, mfr_products, by='COMPANY_NAME') %>%
  mutate(percent_mfr_products_in_shortage = round(n_shortages / n_products * 100), 
         percent_shortages_of_overall = round(n_shortages / sum(n_shortages) * 100), 
         percent_products_of_overall = round(n_products / sum(n_products) * 100)) %>%
    top_n(TOP_N_MFR_MARKETED, wt = n_shortages)
  
write_csv(mfr_percent_shortages, 'figure_9-mfr_product_shortages.csv')

figure_9 = mfr_percent_shortages %>%
  ggplot(aes(x = percent_products_of_overall, y = percent_shortages_of_overall, label = COMPANY_NAME)) + 
    theme_minimal() + 
    geom_point() + 
    geom_abline(slope = 1, intercept = 0, linetype=2, alpha=0.2) + 
    geom_text_repel(size = 2, segment.alpha = 0) +
    coord_cartesian(xlim = c(0, 11), ylim=c(0,17)) + 
    labs(title = paste("Top",TOP_N_MFR_MARKETED, "Drugs Producers by Number of Shortages, Canada, 2013–16"),
         subtitle = paste("Over the analysis period", ANALYSIS_PERIOD_START, "to", ANALYSIS_PERIOD_END),
         caption = paste("Source: DS, DPD   Updated:", now()),
         x = "% of marketed drug products", 
         y = "% of shortages")
ggsave('figure_9-mfr_product_shortages.png', figure_9)
figure_9
```

### Table 3 - Oral Drug shortages by Company

```{r table_3_pre}
oral_shortages = drugshortages_ca_manual_raw %>%
  mutate(start_date = as.Date(start_date)) %>%
  filter(between(start_date, ANALYSIS_PERIOD_START, ANALYSIS_PERIOD_END), 
         route_of_administration == 'Oral')

total_oral_shortages = nrow(oral_shortages)

table_3 = oral_shortages %>%
  group_by(company_name) %>%
  summarize(n_shortages = n(), n_products = n_distinct(din)) %>%
  ungroup() %>%
  arrange(desc(n_shortages)) %>%
  mutate(percent_total_shortages = round(n_shortages/total_oral_shortages * 100,2), 
         cum_percent_shortages = cumsum(n_shortages)/total_oral_shortages*100) %>%
  filter(cum_percent_shortages <= 90) %>%
  bind_rows (
    summarize(.,
      company_name = 'TOTAL', 
      n_shortages = sum(n_shortages), 
      n_products = sum(n_products), 
      percent_total_shortages = sum(percent_total_shortages), 
      cum_percent_shortages = sum(cum_percent_shortages)
      )
  ) %>%
  transmute(
    "Company Name" = company_name, 
    "# of Products (DIN)" = n_products, 
    "# of Shortages" = n_shortages,
    "% of Total Shortages" = percent_total_shortages
  ) 
```

Table 3 Twenty Generic+ and Innovator++ Companies Reporting 90% of `r total_oral_shortages` ‘Oral’ Drug Shortages (April 1st 2013 – March 31st 2016) (Source: www.drugshortage.ca, launched in 2012)

```{r table_3}
knitr::kable(table_3)
```

### Table 4 - Injectable Drug shortages by Company
```{r table_4_pre}
inject_shortages = drugshortages_ca_manual_raw %>%
  mutate(start_date = as.Date(start_date)) %>%
  filter(between(start_date, ANALYSIS_PERIOD_START, ANALYSIS_PERIOD_END), 
         route_of_administration == 'Injectable')

total_inject_shortages = nrow(inject_shortages)

table_4 = inject_shortages %>%
  group_by(company_name) %>%
  summarize(n_shortages = n(), n_products = n_distinct(din)) %>%
  ungroup() %>%
  arrange(desc(n_shortages)) %>%
  mutate(percent_total_shortages = round(n_shortages/total_inject_shortages * 100,2), 
         cum_percent_shortages = cumsum(n_shortages)/total_inject_shortages*100) %>%
  filter(cum_percent_shortages <= 90) %>%
  bind_rows (
    summarize(.,
      company_name = 'TOTAL', 
      n_shortages = sum(n_shortages), 
      n_products = sum(n_products), 
      percent_total_shortages = sum(percent_total_shortages), 
      cum_percent_shortages = sum(cum_percent_shortages)
      )
  ) %>%
  transmute(
    "Company Name" = company_name, 
    "# of Products (DIN)" = n_products, 
    "# of Shortages" = n_shortages,
    "% of Total Shortages" = percent_total_shortages
  ) 
```

Table 4 Twenty Generic+ and Innovator++ Companies Reporting 90% of `r total_inject_shortages` ‘Injectable’ Drug Shortages (April 1st 2013 – March 31st 2016) (Source: www.drugshortage.ca, launched in 2012)

```{r table_4}
knitr::kable(table_4)
```


Drug Products
====================

Row {data-width=300}
-------------------- 

### Background
The Health Canada [Drug Product Database](https://www.canada.ca/en/health-canada/services/drugs-health-products/drug-products/drug-product-database.html) provides historical records of drug products (and provides [extracts](https://www.canada.ca/en/health-canada/services/drugs-health-products/drug-products/drug-product-database/what-data-extract-drug-product-database.html))

For the following charts, we only consider prescription drugs for humans. 

The DPD contains records tracking the change in a DINs status over time (in the `QRYM_STATUS`) table. Drug products are given a status of either: `APPROVED`, `MARKETED`, `CANCELLED POST-MARKET`, `CANCELLED PRE-MARKET`, or `DORMANT`.

The following chart shows, for each year, the number of drug products categorized by their status in that year. Drug product status in a year is taken from the most recent QRYM_STATUS entry to the year.

Row
------------

### DIN Status

```{r din-status-by-year, fig.width=10}
ggplot(drug_yearly_status, aes(x = year.date, fill=status)) + 
  geom_bar() +
  theme_minimal() +
  labs(title = "DIN Status by Year",
       subtitle = "Status determined by most recent status change", 
       caption = paste('Source: Health Canada Drug Product Database. Updated:', now()), 
       x = "Year", y = "DIN Count")
```

### Figure 8: Market Share By MFR

```{r din-manufacturers}
MFR_MIN_PERCENT = 3
MFR_START_YEAR = floor_date(ANALYSIS_PERIOD_START, unit = "year")
MFR_END_YEAR = floor_date(ANALYSIS_PERIOD_END, unit = "year")

active_drugs_10yrs = drug_yearly_status %>%
  filter(between(year.date, MFR_START_YEAR, MFR_END_YEAR), status %in% c('MARKETED')) %>%
  group_by(DRUG_CODE) %>%
  summarise() %>%
  inner_join(dpd_comp, by="DRUG_CODE") %>%
  group_by(COMPANY_NAME) %>%
  summarize(n = n()) %>%
  mutate(percent = n/sum(n) * 100, 
         company_name = if_else(percent >= MFR_MIN_PERCENT, COMPANY_NAME, 'other')) %>%
  group_by(company_name) %>%
  summarize(percent = sum(percent), n_companies = n()) %>%
  mutate(company_name = if_else(company_name == 'other', 
                                paste0('Other (', n_companies, ' manufacturers, ', round(percent), '%)'), 
                                paste0(company_name, ' (', round(percent), '%)')))

active_drugs_10yrs$company_name = fct_shift(fct_reorder(active_drugs_10yrs$company_name, -active_drugs_10yrs$percent), n=1)

write_csv(active_drugs_10yrs, 'figure_8-company_market_share.csv')

figure_8 = active_drugs_10yrs %>%
  ggplot(aes(x = "", y = percent, fill = company_name)) + 
    #theme(legend.position = "none") + 
    geom_bar(stat = 'identity') + 
    theme_void() + 
    scale_fill_discrete(guide = guide_legend(title = "Manufacturer")) + 
    coord_polar(theta = 'y', direction = -1) + 
    labs(title = "Market Share by Major Drugs Producer, Canada, 2013–16",
         subtitle = paste("Only manufacturers of >= %",MFR_MIN_PERCENT," of DINs listed explicitly", sep=""), 
         caption = paste("Source: DPD, Updated: ", now()), 
         x = '', y = '')
ggsave('figure_8-company_market_share.png', plot = figure_8)
figure_8
```

Innovator vs. Generics
===========================

Row {data-width=300}
--------------------

### Background 

In this section we look at innovator vs. generic drug products historically. To
determine whether a drug product is an innovator or generic we make the
assumption innovator drug products are always created before generic drug
products. Therefore in a group of drug products with the same active
ingredients, the earliest manufacturer is the innovator and all their drug
products in the group are marked as such. All other drug products, are marked as
generic.

[From the DPD page on Terminology](https://www.canada.ca/en/health-canada/services/drugs-health-products/drug-products/drug-product-database/terminology.html): 

"The AIG [Active Ingredient Group] number is a 10 digit number that identifies products that have the same active   ingredient(s) and ingredient strength(s). The AIG is comprised of three portions:

  - the first portion (2 digits) identifies the number of active ingredients
  - the second portion(5 digits) identifies the unique groups of active ingredients(s);
  - the last portion (3 digits) identifies the active ingredient group strength. The   strength group has a tolerance of -2% to +10%."

Innovator/Generic status is determined within groups determined by the second portion of the AIG only so that drug strength is ignored.

The following chart shows only the active drug products (`MARKETED`) over time. 

```{r innovator-setup, include = FALSE}
# A list of innovator drug companies per ingredient group (IG)
innovator_drugs = drug_product %>%
  select(DRUG_CODE, IG) %>%
  inner_join(dpd_status, by = 'DRUG_CODE') %>%
  inner_join(dpd_comp, by = 'DRUG_CODE') %>%
  group_by(IG) %>%
  arrange(HISTORY_DATE, .by_group = T) %>%
  summarise(COMPANY_CODE = first(COMPANY_CODE))
  
# All drug products with the Innovator/Generic status
drug_product_innovator = 
  drug_product %>%
  inner_join(dpd_comp, by='DRUG_CODE') %>%
  inner_join(innovator_drugs, by = 'IG', suffix = c(".p", ".innovator")) %>%
  mutate(is_innovator = 
           if_else(COMPANY_CODE.p == COMPANY_CODE.innovator, 'Innovator', 'Generic')) %>%
  select(DRUG_CODE, DRUG_IDENTIFICATION_NUMBER, AI_GROUP_NO, IG, is_innovator)

# Yearly counts of active (MARKETED/APPROVED) drugs categorized by Innovator/Generic status
drug_yearly_status_innovator = 
  inner_join(drug_product_innovator, drug_yearly_status, by='DRUG_CODE') %>% 
  mutate(is_innovator = as.factor(is_innovator))
```


Row
--------------
### Figure 10: Drug Products (DINs) by Year

```{r}
marketed_status = drug_yearly_status_innovator %>%
  filter(status %in% c('MARKETED')) 

# rearrange to export results for the plot
write_csv(marketed_status %>%
  group_by(year.date, is_innovator) %>%
  summarize(n = n()), 'figure_10-marketed-DINs-over-time.csv')

figure_10 = marketed_status %>%
  ggplot(aes(x = year.date, fill = is_innovator)) + 
    geom_bar() + 
    theme_minimal() + 
    theme(legend.position = 'bottom') +
    scale_fill_discrete(guide = guide_legend(title = "Innovator Status")) +
    labs(title = "Drug Products (DINs) Available Each Year, Canada, 1970–2017", 
         subtitle = 'Innovator drugs determined per IG group by company first to market',
         caption = paste('Source: Health Canada Drug Product Database. Updated:', now()), 
         x = "Year", 
         y = "DIN Count") 
ggsave('figure_10-marketed-DINs-over-time.png', plot = figure_10)
figure_10
```

Supplementary Figures
=====================

Row {.tabset}
--------------

### Shortages by Reasons
```{r shortages-reasons, fig.width=8}

drugshortagescanada_ca %>%
  mutate(year = year(start_date)) %>%
  filter(year == 2017) %>%
  mutate(month = month(start_date, label = T, abbr = TRUE)) %>%
  ggplot(aes(x = month, fill = Reason)) + 
    geom_bar() + 
    labs(title = "Reasons for shortages in drugshortagescanada.ca", 
         subtitle = "")
```

### Shortages by Duration

```{r shortages-by-duration, fig.width=8}
#bind_rows(drugshortagescanada_ca, drugshortages_ca)
drugshortagescanada_ca %>%
  mutate(duration = end_date - start_date) %>%
  filter(duration < 365 & duration > 0) %>%
  ggplot(aes(x=duration, fill=Reason)) + 
    geom_histogram(binwidth = 30) + 
    theme_minimal() + 
    facet_grid( . ~ source) + 
    labs(title = "Duration of drug shortages", 
         x = "Duration (days)",
         y = "Number of shortagse")
```

### Shortages by Duration/Resolution

```{r duration-resolution}

#bind_rows(drugshortagescanada_ca, drugshortages_ca)
drugshortages_ca %>%
  mutate(duration = end_date - start_date, 
         resolution = fct_relevel(as_factor(resolution), "Discontinued", "Resolved", "Current")) %>%
  filter(duration < 365 & duration > 0) %>%
  ggplot(aes(x=duration, fill=resolution)) + 
    geom_histogram(binwidth = 30) + 
    theme_minimal() + 
    facet_grid( . ~ source) + 
    labs(title = "Duration of drug shortages", 
         subtitle = "Only shortages <1 year are shown",
         x = "Duration (days)",
         y = "Number of shortagse")
```

### Shortages Per Manufacturer
```{r manufacturers-by-num-shortages}
MFR_MIN_PERCENT = 5

mfr_shortages = drugshortages_ca %>% 
  filter(between(start_date, ANALYSIS_PERIOD_START, ANALYSIS_PERIOD_END)) %>%
  inner_join(drug_product, by=c('DIN' = 'DRUG_IDENTIFICATION_NUMBER')) %>%
  inner_join(dpd_comp, by='DRUG_CODE') %>%
  group_by(COMPANY_NAME) %>%
  summarize(n_shortages = n()) %>%
  mutate(percent = n_shortages/sum(n_shortages) * 100, 
         company_name = if_else(percent >= MFR_MIN_PERCENT, COMPANY_NAME, 'other')) %>%
  group_by(company_name) %>%
  summarize(percent = sum(percent), n_companies = n(), n_shortages = sum(n_shortages)) %>%
  mutate(company_name = if_else(company_name == 'other', 
                                paste0('Other (', n_companies, ' manufacturers, ', n_shortages, 
                                       ' shortages, ', round(percent), '%)'), 
                                paste0(company_name, ' (', n_shortages, ' shortages, ', 
                                       round(percent), '%)')))

mfr_shortages$company_name = fct_shift(fct_reorder(mfr_shortages$company_name, -mfr_shortages$percent), n=1)

mfr_shortages %>%
  ggplot(aes(x = "", y = percent, fill = company_name)) + 
    geom_bar(stat = 'identity') + 
    theme_void() + 
    scale_fill_discrete(guide = guide_legend(title = "Manufacturer")) + 
    coord_polar(theta = 'y', direction = -1) + 
    labs(title = paste0("% of shortages during analysis period (", 
                       ANALYSIS_PERIOD_START, " to ", ANALYSIS_PERIOD_END, ')'), 
         caption = paste("Source: DS, Updated: ", now()), 
         x = '', y = '')
```

### Shortages by Route

```{r shortages_by_route}
drugshortages_ca_manual_raw %>%
  mutate(start_date = as.Date(start_date)) %>%
  filter(between(start_date, ANALYSIS_PERIOD_START, ANALYSIS_PERIOD_END), 
         route_of_administration %in% c('Oral', 'Injectable')) %>%
  group_by(company_type, route_of_administration) %>%
  summarise(Products = n_distinct(product_name), Shortages = n()) %>%
  gather(Measure, Count, -company_type, -route_of_administration) %>%
  ggplot(aes(x = company_type, y = Count, fill=Measure)) + 
    geom_bar(stat='identity', position='dodge') + 
    theme_minimal() +
    theme(legend.position = "bottom") + 
    scale_fill_discrete(guide = guide_legend(title = "")) + 
    facet_wrap(~route_of_administration) + 
    labs(x = "")
```
